---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# gestalt

Every non-primitive function in R chains together other functions. The _gestalt_
package enables you to naturally express this using the `%>>>%` operator. It
adopts the semantics of the [magrittr](https://magrittr.tidyverse.org) `%>%`, in
addition to supporting tidyverse-style
[quasiquotation](https://rlang.r-lib.org/reference/quasiquotation.html).

Use it
to improve the modularity, reusability and comprehensibility of your functions.

```{r}
library(gestalt)

model <- mpg ~ wt

r2_cyl <- 
  group: split(.$cyl) %>>>%
  lapply(function(data) lm(!!model, data)) %>>>% 
  summary: (
    lapply(summary) %>>>%
    stat: sapply(`[[`, "r.squared")
  )

r2_cyl(mtcars)

(split(.$gear) %>>>% r2_cyl[-1])(mtcars)

r2 <- r2_cyl
r2$group <- function(x, grp) split(x, x[[grp]])
r2(mtcars, "gear")

residuals <- r2
residuals$summary$stat <- function(s) lapply(s, `[[`, "residuals")
residuals(mtcars, "cyl")[["6"]]
```

## Installation

Install the development version from GitHub:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("egnha/gestalt")
```
## License

MIT Copyright © 2018 [Eugene Ha](https://github.com/egnha)
